---
title: "Domain decomposition"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Domain decomposition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(hydroloom)
library(dplyr)

local <- (Sys.getenv("BUILD_VIGNETTES") == "TRUE")

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center",
  eval = local
)

oldoption <- options(scipen = 9999)
```

# The trunk--compact relationship

The domain decomposition model implemented in hydroloom splits a network
into two types of domains: trunk domains and compact (leaf or stem) domains.

A trunk is the set of main rivers that connect and flow through compact domains. The 
incremental flowlines and catchments that make up a trunk belong to the
compact domain that the trunk flows through. The accumulated values
that reflect the broader network connectivity of the trunk belongs to
the trunk domain only. 

A compact domain is a catchment aggregate (HY_Features): its drainage
area is the full contributing area to the trunk segment that passes
through it, including the trunk's own incremental catchments along that
segment. So the catchments along the trunk's mainstem are, by area, part
of the surrounding compact domain. The trunk itself is not a catchment. It
carries accumulated values from one compact domain to the next.

The trunk carries accumulated values along its incremental catchments.
A compact domain accumulates internally -- every tributary inside it drains to
the compact domain's outlet for headwater compact domains, or to many outlets 
along the trunk. The compact-to-trunk outlets hold accumulated values
until recomposition. At recomposition, the compact domain passes its outlet
values to the trunk, and the trunk continues the basin's downstream
accumulation from there.

In practice, this split has three consequences:

- A compact domain's catchment network is self-contained, so per-domain
  processing can be run in parallel. Similarly, one compact domain can be edited
  or re-attributed without invalidating any of the others.
- Recomposition is a single pass through the domain graph. Trunk-level
  accumulation regenerates from per-compact outlet values; catchment-level
  data does not need to be revisited.
- Editing and updating decouple. A new modeled result, a corrected
  geometry, or a re-derived attribute lands in one compact domain at a time, and
  the trunk catches up by replaying recomposition.

# Use case

A catchment network that covers a large area contains many features with
complex upstream-downstream dependencies. `decompose_network()` partitions
such a network into self-contained sub-networks -- called domains -- so that
each domain can be resolved independently. 

**"Compact"** domains make up the entire contributing area that flows to a trunk 
domain but do not include the connectivity of the trunk domain. **"Trunk"** domains 
span an entire drainage basin, do not include local contributing area information but 
do include connectivity of major rivers.  

# Decomposed and recomposed modes

A compact domain's `catchments` slot holds two kinds of rows: the lateral
tributary rows and the trunk-mainstem rows that flow through the
compact domain's segment. On the trunk-mainstem rows, the `toid` is dropped to
the outlet sentinel returned by `get_outlet_value()`, so each becomes a
local outlet of its own little contributing sub-basin. This is the
*decomposed* mode -- the default form `decompose_network()` returns.
Trunk-mainstem membership is recoverable from the parent trunk domain's
`catchments$id`; no marker column on the compact domain is needed.

In decomposed mode a single `accumulate_downstream()` call on a
compact domain's catchments produces, for every trunk-mainstem row, the
locally-incremental drainage area (or any other accumulable) belonging
there: the trunk row's own incremental plus the lateral tributaries that
drain into it. Per-compact processing is independent and runs in
parallel.

To switch to *recomposed* mode, restore the trunk-mainstem rows' toids
from `source_network`:

```r
restored <- compact$catchments
src_lookup <- setNames(source_network$toid, source_network$id)
sentinel <- get_outlet_value(restored)
detoid_idx <- which(restored$toid == sentinel)
restored$toid[detoid_idx] <- src_lookup[as.character(restored$id[detoid_idx])]
```

The compact domain's segment is now a connected sub-basin again; standard
hydroloom operations (model coupling, hydraulic routing, end-to-end
visualization) work unchanged. Switching back is a one-column
assignment.

# Functional characteristics

The decomposition operates on drainage basins. A drainage basin
(an HY_Features concept) is the total upstream area draining to an
outlet, with a single mainstem connecting its headwater to that outlet.
The input network may contain one or more drainage basins, each composed
of incremental catchments connected in an upstream-downstream topology.

For each drainage basin, `decompose_network()` identifies the trunk
catchments -- the basin's mainstem (outlet levelpath) by default, or
the catchments above a threshold when `trunk_threshold` is set -- and
divides the trunk into segments at confluences. Each segment becomes a
compact domain whose catchments slot holds the lateral tributaries
draining into the segment together with the segment's trunk-mainstem
rows in their decomposed (detoid'd) form. Even segments with no
lateral tributaries
become compact domains -- the trunk-mainstem rows alone constitute the
compact domain's contributing area.

A trunk domain is built alongside the compact domains. It carries the
mainstem catchments with their toids intact, so the connected mainstem
is addressable for drawing, naming, or any operation that wants the
end-to-end flowpath. The trunk is the same set of catchments that
appear, in their decomposed form, across the compact domains. The duplication
keeps the mainstem queryable as a single object without forcing readers
to reassemble it from compact domains.

A compact domain corresponds to an HY_Features catchment aggregate.
Its drainage area is the full contributing area to its trunk segment.
The decomposition does not materialize catchment aggregate objects
directly; the alignment is by construction, with the compact domain's
catchments slot holding the rows the aggregate would.

The trunk domain discharges at the outlet of the drainage basin. The
inter-domain graph records flow between compact domains through synthetic
nexus identifiers, with one edge per inter-segment handoff.

# Separation of concerns

The decomposition separates three concerns:

1. **Per-domain computation.** Each domain's catchment network is
   self-contained. Processing a domain requires no information from any
   other domain beyond what arrives at inlet nexuses.
2. **Inter-domain connectivity.** The domain graph captures which domains
   flow into which, and at what nexus -- a coarser-resolution network
   that can be traversed independently of catchment-level detail.
3. **Global ordering.** Each domain carries a `topo_sort_offset` that
   anchors its internal topological sort within a global sequence, so
   cross-domain ordering is recoverable after independent resolution.

# Data model

The decomposition produces a `domain_decomposition` object with six slots:

- **`domains`** -- named list of `hy_domain` objects, one per sub-network.
- **`domain_graph`** -- inter-domain flow edges (`id`, `toid`, `nexus_id`,
  `relation_type`).
- **`overrides`** -- non-dendritic inter-domain transfer table, or `NULL`.
- **`catchment_domain_index`** -- named character vector mapping each
  catchment id to its domain id.
- **`nexus_registry`** -- synthetic nexus identifiers and the domains they
  connect.
- **`source_network`** -- the original enriched input network.

## `hy_domain`

A `hy_domain` bundles a domain's catchment network with the connectivity
metadata that recomposition needs:

- `domain_id` -- unique identifier.
- `domain_type` -- `"trunk"` or `"compact"`.
- `outlet_nexus_id` -- primary outlet nexus (trunk domains have one; compact
  domains may discharge at additional nexuses recorded in the domain graph).
- `inlet_nexus_ids` -- nexuses where compact domains inject lateral inflow
  (empty for compact domains themselves).
- `trunk_domain_id` -- the receiving trunk (for compact domains), or `NA` for trunks.
- `containing_domain_id` -- the enclosing domain for endorheic basins,
  or `NA`.
- `catchments` -- hydroloom object carrying the domain's catchment network.
  Compact catchments include both lateral tributary rows and the segment's
  trunk-mainstem rows; the trunk-mainstem rows have a `toid` set to
  `get_outlet_value()`. To tell the two apart, intersect the compact domain's
  `catchments$id` with the parent trunk domain's `catchments$id`.
- `topo_sort_offset` -- global topo_sort base for cross-domain ordering.

Trunk domains require `hy_leveled` catchments. Compact domains accept
`hy_topo`, `hy_leveled`, or `hy_flownetwork`.

# S3 classes

The decomposition builds on the hydroloom S3 class hierarchy described
in `vignette("hydroloom")`. The relevant classes are:

- **`hy_topo`** -- self-referencing edge list (`id`, `toid`; unique `id`).
- **`hy_leveled`** -- subclass of `hy_topo` enriched with `topo_sort`,
  `levelpath`, and `levelpath_outlet_id`. Required input for
  `decompose_network()`, because the partition uses `levelpath` to identify
  trunk membership and `topo_sort` to establish processing order.
- **`hy_flownetwork`** -- non-dendritic edge list (`id`, `toid`; non-unique
  `id` allowed). Compact domains may hold `hy_flownetwork` catchments to
  preserve internal divergences.

# Running a decomposition

The input must be enriched to `hy_leveled` before decomposition.
`add_toids()` and `add_levelpaths()` handle this.

```{r decompose}
library(hydroloom)

g <- sf::read_sf(system.file("extdata/new_hope.gpkg", package = "hydroloom"))

src <- hy(g) |>
  add_toids(return_dendritic = TRUE) |>
  add_levelpaths(name_attribute = "GNIS_ID",
    weight_attribute = "arbolate_sum")

d <- decompose_network(src)
```

The returned `domain_decomposition` prints a summary of the partition:

```{r print}
d
```

## Trunk selection

`decompose_network()` determines trunk membership through one of three paths:

- **Default** (both `trunk_threshold` and `trunk_levelpaths` are `NULL`) --
  the trunk contains catchments on the basin's terminal-outlet levelpath.
- **Threshold-based** -- when `trunk_threshold` is supplied, a basin's outlet
  metric must exceed the threshold to receive a trunk. Basins below the
  threshold become a single compact domain with no trunk extraction.
- **Explicit** -- when `trunk_levelpaths` is supplied, those levelpaths
  (plus the outlet levelpath) form the trunk.

```{r threshold}
d_thresh <- decompose_network(src, trunk_threshold = 100)

d_thresh
```

The map below colors each flowline by the compact domain it belongs to and overlays the
trunk in black. Every flowline belongs to some compact domain (the partition
covers the network); the trunk overlay highlights the rows that are
*also* in a trunk domain -- the connected mainstem that the compact domains
turn into local outlets.

```{r plot-domains, fig.width = 7, fig.height = 7}
# Build a domain_id column on the source flowlines.
domain_idx <- d_thresh$catchment_domain_index
plot_net <- d_thresh$source_network
plot_net$domain_id <- unname(domain_idx[as.character(plot_net$id)])

# Pull all trunk-mainstem ids from every trunk domain.
trunk_ids <- unlist(lapply(
  Filter(\(x) x$domain_type == "trunk", d_thresh$domains),
  \(x) as.character(x$catchments$id)),
  use.names = FALSE)

compact_ids <- sort(unique(plot_net$domain_id))

set.seed(42)
pal <- sample(grDevices::hcl.colors(length(compact_ids), palette = "Set 2"))
names(pal) <- compact_ids

par(mar = c(0, 0, 1, 0))
plot(sf::st_geometry(plot_net),
  col = pal[plot_net$domain_id],
  lwd = 2, main = "Domain decomposition (trunk_threshold = 100)")
plot(sf::st_geometry(plot_net[as.character(plot_net$id) %in% trunk_ids, ]),
  col = "black", lwd = 2, add = TRUE)
legend("topright", legend = c("trunk overlay",
  "compact (each color = one domain)"),
  col = c("black", "grey60"), lwd = 2, bty = "n", cex = 0.8)
```

## Manual domain breaks

The automatic segmentation splits the trunk at confluences -- every point
where a tributary enters the trunk starts a new segment. A large threshold
pulls more levelpaths into the trunk and produces fewer, larger compact
domains. The `domain_breaks` parameter gives direct control over where the
trunk is segmented, independent of the confluence structure.

In practice, this is useful when domain boundaries need to align with
modeling sub-basins, gauge locations, or administrative units rather than
the topology's natural confluence points. The approach is to set a high
threshold (so the trunk absorbs most of the network) and then pass specific
catchment ids as break points.

```{r manual-breaks}
# A high threshold pulls many levelpaths into the trunk. Without
# domain_breaks the trunk is segmented at every confluence, which
# may produce more compact domains than desired.
d_high <- decompose_network(src, trunk_threshold = 25)
d_high

# Identify trunk catchments where we want to force segment boundaries.
# Here we pick two catchments roughly in the middle of the trunk.
trunk_dom <- Filter(\(x) x$domain_type == "trunk", d_high$domains)[[1]]
trunk_catch <- trunk_dom$catchments

# Choose break points at the 1/3 and 2/3 marks along the trunk.
break_ids <- trunk_catch$id[
  c(floor(nrow(trunk_catch) / 3), floor(2 * nrow(trunk_catch) / 3))]

d_manual <- decompose_network(src,
  trunk_threshold = 25,
  domain_breaks = break_ids)
d_manual
```

The map below compares the automatic segmentation (left) with the
manual-break segmentation (right). Both use the same trunk threshold; the
difference is where compact domains attach along the trunk.

```{r plot-manual-breaks, fig.width = 10, fig.height = 5}
plot_decomp <- function(d, title) {

  domain_idx <- d$catchment_domain_index
  net <- d$source_network
  net$domain_id <- unname(domain_idx[as.character(net$id)])

  trunk_ids <- unlist(lapply(
    Filter(\(x) x$domain_type == "trunk", d$domains),
    \(x) as.character(x$catchments$id)),
    use.names = FALSE)

  compact_ids <- sort(unique(net$domain_id))

  set.seed(42)
  pal <- sample(grDevices::hcl.colors(
    max(length(compact_ids), 1), palette = "Set 2"))
  names(pal) <- compact_ids

  plot(sf::st_geometry(net),
    col = pal[net$domain_id],
    lwd = 2, main = title)
  plot(sf::st_geometry(net[as.character(net$id) %in% trunk_ids, ]),
    col = "black", lwd = 2, add = TRUE)

}

par(mfrow = c(1, 2), mar = c(0, 0, 2, 0))
plot_decomp(d_high, "Automatic segmentation")
plot_decomp(d_manual, "Manual domain_breaks")
```

## Accessing domains

Individual domains, the domain graph, and the catchment-to-domain mapping are
accessible through the slots of the returned object:

```{r access}
# number and types of domains
types <- vapply(d$domains, \(x) x$domain_type, character(1))
table(types)

# inter-domain flow edges
head(d$domain_graph)

# look up which domain owns a catchment
get_domain_for_catchment(d, d$source_network$id[1:5])
```

```{r teardown, include = FALSE}
options(oldoption)
```
