---
title: "Domain decomposition"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Domain decomposition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(hydroloom)
library(dplyr)

local <- (Sys.getenv("BUILD_VIGNETTES") == "TRUE")

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center",
  eval = local
)

oldoption <- options(scipen = 9999)
```

# Use case

A catchment network that covers a large area contains many features with
complex upstream-downstream dependencies. `decompose_network()` partitions
such a network into self-contained sub-networks -- called domains -- so that
each domain can be resolved independently. Compact domains (tributaries and
headwaters) connect only to their parent trunk domain, never to each other,
which makes them independently computable. Trunk domains carry the primary
flowpath network of a drainage basin and receive lateral inflow from compact
domains at nexus points along their length.

# Functional characteristics

The decomposition operates on drainage basins. A drainage basin
(an HY_Features concept) is the total upstream area draining to an outlet,
with a single mainstem connecting its headwater to that outlet. The input
network may contain one or more drainage basins, each composed of
incremental catchments connected in an upstream-downstream topology.

For each drainage basin, `decompose_network()` extracts the primary flowpath
catchments into a trunk domain and groups the remaining catchments -- the
lateral tributaries -- into compact domains. In the default configuration the
trunk follows the basin's mainstem (outlet levelpath); with a threshold it may
include catchments from multiple levelpaths. The trunk is not itself a drainage
basin (it has lateral inflows along its length), and the compact domains are
tributary sub-networks rather than complete basins.

Compact domains are grouped by trunk segment -- a maximal chain of trunk
catchments between two confluences. All tributaries entering the same
stretch of trunk merge into a single compact domain, which may connect to
the trunk at multiple nexus points along that stretch. This grouping
produces domains of comparable size rather than one domain per tributary,
and because compact domains connect only to their parent trunk (never to
each other), they remain independently computable.

The decomposition is organized around the HY_Features catchment aggregate
concept. A compact domain corresponds to a catchment aggregate -- the
segment of trunk that flows through it is the flowpath of the aggregate that
the compact represents. The decomposition does not materialize catchment
aggregate objects directly; the alignment is conceptual, as expected for a
conceptual model like HY_Features.

The trunk domain discharges at the outlet of the drainage basin. A domain
graph records inter-domain flow connections through synthetic nexus
identifiers, with one edge per nexus point where a compact meets the trunk.

# Separation of concerns

The decomposition separates three concerns:

1. **Per-domain computation.** Each domain's catchment network is
   self-contained. Processing a domain requires no information from any
   other domain beyond what arrives at inlet nexuses.
2. **Inter-domain connectivity.** The domain graph captures which domains
   flow into which, and at what nexus -- a coarser-resolution network
   that can be traversed independently of catchment-level detail.
3. **Global ordering.** Each domain carries a `topo_sort_offset` that
   anchors its internal topological sort within a global sequence, so
   cross-domain ordering is recoverable after independent resolution.

# Data model

The decomposition produces a `domain_decomposition` object with six slots:

- **`domains`** -- named list of `hy_domain` objects, one per sub-network.
- **`domain_graph`** -- inter-domain flow edges (`id`, `toid`, `nexus_id`,
  `relation_type`).
- **`overrides`** -- non-dendritic inter-domain transfer table, or `NULL`.
- **`catchment_domain_index`** -- named character vector mapping each
  catchment id to its domain id.
- **`nexus_registry`** -- synthetic nexus identifiers and the domains they
  connect.
- **`source_network`** -- the original enriched input network.

## `hy_domain`

A `hy_domain` bundles a domain's catchment network with the connectivity
metadata that recomposition needs:

- `domain_id` -- unique identifier.
- `domain_type` -- `"trunk"` or `"compact"`.
- `outlet_nexus_id` -- primary outlet nexus (trunk domains have one; compact
  domains may discharge at additional nexuses recorded in the domain graph).
- `inlet_nexus_ids` -- nexuses where compact domains inject lateral inflow
  (empty for compact domains themselves).
- `trunk_domain_id` -- the receiving trunk (for compacts), or `NA` for trunks.
- `containing_domain_id` -- the enclosing domain for endorheic basins,
  or `NA`.
- `catchments` -- hydroloom object carrying the domain's catchment network.
- `topo_sort_offset` -- global topo_sort base for cross-domain ordering.

Trunk domains require `hy_leveled` catchments. Compact domains accept
`hy_topo`, `hy_leveled`, or `hy_flownetwork`.

# S3 classes

The decomposition builds on the hydroloom S3 class hierarchy described
in `vignette("hydroloom")`. The relevant classes are:

- **`hy_topo`** -- self-referencing edge list (`id`, `toid`; unique `id`).
- **`hy_leveled`** -- subclass of `hy_topo` enriched with `topo_sort`,
  `levelpath`, and `levelpath_outlet_id`. Required input for
  `decompose_network()`, because the partition uses `levelpath` to identify
  trunk membership and `topo_sort` to establish processing order.
- **`hy_flownetwork`** -- non-dendritic edge list (`id`, `toid`; non-unique
  `id` allowed). Compact domains may hold `hy_flownetwork` catchments to
  preserve internal divergences.

# Running a decomposition

The input must be enriched to `hy_leveled` before decomposition.
`add_toids()` and `add_levelpaths()` handle this.

```{r decompose}
library(hydroloom)

g <- sf::read_sf(system.file("extdata/new_hope.gpkg", package = "hydroloom"))

src <- hy(g) |>
  add_toids(return_dendritic = TRUE) |>
  add_levelpaths(name_attribute = "GNIS_ID",
    weight_attribute = "arbolate_sum")

d <- decompose_network(src)
```

The returned `domain_decomposition` prints a summary of the partition:

```{r print}
d
```

## Trunk selection

`decompose_network()` determines trunk membership through one of three paths:

- **Default** (both `trunk_threshold` and `trunk_levelpaths` are `NULL`) --
  the trunk contains catchments on the basin's terminal-outlet levelpath.
- **Threshold-based** -- when `trunk_threshold` is supplied, a basin's outlet
  metric must exceed the threshold to receive a trunk. Basins below the
  threshold become a single compact domain with no trunk extraction.
- **Explicit** -- when `trunk_levelpaths` is supplied, those levelpaths
  (plus the outlet levelpath) form the trunk.

```{r threshold}
d_thresh <- decompose_network(src, trunk_threshold = 100)

d_thresh
```

The map below colors each flowline by its domain assignment. The trunk domain
(thick black lines) carries the primary flowpath; compact domains (colored)
are the lateral sub-networks grouped along each trunk segment.

```{r plot-domains, fig.width = 7, fig.height = 7}
# build a domain_id column on the source flowlines
domain_idx <- d_thresh$catchment_domain_index
plot_net <- d_thresh$source_network
plot_net$domain_id <- unname(domain_idx[as.character(plot_net$id)])

types <- setNames(
  vapply(d_thresh$domains, \(x) x$domain_type, character(1)),
  vapply(d_thresh$domains, \(x) x$domain_id, character(1))
)

plot_net$domain_type <- unname(types[plot_net$domain_id])

# color compact domains; trunk drawn separately
compact_ids <- unique(plot_net$domain_id[plot_net$domain_type == "compact"])

set.seed(42)
pal <- sample(grDevices::hcl.colors(length(compact_ids), palette = "Set 2"))
names(pal) <- compact_ids

par(mar = c(0, 0, 1, 0))
plot(sf::st_geometry(plot_net[plot_net$domain_type == "compact", ]),
  col = pal[plot_net$domain_id[plot_net$domain_type == "compact"]],
  lwd = 2, main = "Domain decomposition (trunk_threshold = 100)")
plot(sf::st_geometry(plot_net[plot_net$domain_type == "trunk", ]),
  col = "black", lwd = 2, add = TRUE)
legend("topright", legend = c("trunk", "compact (each color = one domain)"),
  col = c("black", "grey60"), lwd = 2, bty = "n", cex = 0.8)
```

## Manual domain breaks

The automatic segmentation splits the trunk at confluences -- every point
where a tributary enters the trunk starts a new segment. A large threshold
pulls more levelpaths into the trunk and produces fewer, larger compact
domains. The `domain_breaks` parameter gives direct control over where the
trunk is segmented, independent of the confluence structure.

In practice, this is useful when domain boundaries need to align with
modeling sub-basins, gauge locations, or administrative units rather than
the topology's natural confluence points. The approach is to set a high
threshold (so the trunk absorbs most of the network) and then pass specific
catchment ids as break points.

```{r manual-breaks}
# A high threshold pulls many levelpaths into the trunk. Without
# domain_breaks the trunk is segmented at every confluence, which
# may produce more compact domains than desired.
d_high <- decompose_network(src, trunk_threshold = 25)
d_high

# Identify trunk catchments where we want to force segment boundaries.
# Here we pick two catchments roughly in the middle of the trunk.
trunk_dom <- Filter(\(x) x$domain_type == "trunk", d_high$domains)[[1]]
trunk_catch <- trunk_dom$catchments

# Choose break points at the 1/3 and 2/3 marks along the trunk.
break_ids <- trunk_catch$id[
  c(floor(nrow(trunk_catch) / 3), floor(2 * nrow(trunk_catch) / 3))]

d_manual <- decompose_network(src,
  trunk_threshold = 25,
  domain_breaks = break_ids)
d_manual
```

The map below compares the automatic segmentation (left) with the
manual-break segmentation (right). Both use the same trunk threshold; the
difference is where compact domains attach along the trunk.

```{r plot-manual-breaks, fig.width = 10, fig.height = 5}
plot_decomp <- function(d, title) {

  domain_idx <- d$catchment_domain_index
  net <- d$source_network
  net$domain_id <- unname(domain_idx[as.character(net$id)])

  types <- setNames(
    vapply(d$domains, \(x) x$domain_type, character(1)),
    vapply(d$domains, \(x) x$domain_id, character(1)))
  net$domain_type <- unname(types[net$domain_id])

  compact_ids <- unique(net$domain_id[net$domain_type == "compact"])

  set.seed(42)
  pal <- sample(grDevices::hcl.colors(
    max(length(compact_ids), 1), palette = "Set 2"))
  names(pal) <- compact_ids

  plot(sf::st_geometry(net[net$domain_type == "compact", ]),
    col = pal[net$domain_id[net$domain_type == "compact"]],
    lwd = 2, main = title)
  plot(sf::st_geometry(net[net$domain_type == "trunk", ]),
    col = "black", lwd = 2, add = TRUE)

}

par(mfrow = c(1, 2), mar = c(0, 0, 2, 0))
plot_decomp(d_high, "Automatic segmentation")
plot_decomp(d_manual, "Manual domain_breaks")
```

## Accessing domains

Individual domains, the domain graph, and the catchment-to-domain mapping are
accessible through the slots of the returned object:

```{r access}
# number and types of domains
types <- vapply(d$domains, \(x) x$domain_type, character(1))
table(types)

# inter-domain flow edges
head(d$domain_graph)

# look up which domain owns a catchment
get_domain_for_catchment(d, d$source_network$id[1:5])
```

```{r teardown, include = FALSE}
options(oldoption)
```
