---
title: "Domain decomposition"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Domain decomposition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(hydroloom)
library(dplyr)

local <- (Sys.getenv("BUILD_VIGNETTES") == "TRUE")

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  fig.align = "center",
  eval = local
)

oldoption <- options(scipen = 9999)
```

# Domains and the extensive connectivity

`decompose_network()` partitions a hydrologic network into independent
sub-basins called *domains* and stores each basin's regionally-connecting
flowlines separately as the basin's *extensive network* — the features
that realize its *extensive connectivity*. Every `hy_domain` is
functionally the same: a self-contained piece of one drainage basin,
holding the segment's lateral tributaries together with the extensive
network rows that flow through the segment.

The extensive network is the basin's connecting flowlines with their
`toid`s intact — a `hy_leveled` view, addressable end-to-end for
drawing, naming, or any operation that wants the full network. The
extensive network rows that appear in the overlay also appear inside
the domains, but with their `toid`s set to the reserved outlet value
returned by `get_outlet_value()`. Each becomes a local outlet of its
own contributing sub-basin.

The duplication is deliberate. With the same rows held both ways, a
domain's accumulation runs as a single pass over its catchments and
the basin's extensive network stays queryable as one object. Editing
or re-attributing a single domain does not invalidate any other.

# Use case

A catchment network covering a large area carries many features with upstream-downstream dependencies. `decompose_network()`
splits such a network into self-contained sub-basins so that each
domain can be processed independently, then recomposed into a single
result.

In practice, this gives three things:

- A domain's catchment network is self-contained, so per-domain
  processing can be run in parallel.
- Recomposition is a single pass through the inter-domain graph
  derived from the nexus registry. Per-domain results combine through
  the basin's extensive network to produce basin-wide answer.
- Editing is decoupled. A new modeled result, a corrected geometry, or a
  re-derived attribute lands in one domain at a time, and the basin
  catches up by replaying recomposition.

# Decomposed and recomposed modes

A domain's `catchments` slot holds two kinds of rows: the lateral
tributary rows and the extensive network rows that flow through the
domain's segment. On the extensive network rows, the `toid` is set to
the reserved outlet value returned by `get_outlet_value()`, so each
becomes a local outlet of its own contributing sub-basin. This is the
*decomposed* mode — the default form `decompose_network()` returns.
Extensive network membership inside a domain is recoverable by
intersecting the domain's `catchments$id` with the basin's extensive
network overlay's `id`.

In decomposed mode a single `accumulate_downstream()` call on a
domain's catchments produces, for every extensive network row, the
locally-incremental drainage area (or any other accumulable) belonging
there: the row's own incremental plus the lateral tributaries that
drain into it. Per-domain processing is independent and runs in
parallel.

To switch to *recomposed* mode, restore the extensive network rows'
toids from `source_network`:

```r
restored <- domain$catchments
src_lookup <- setNames(source_network$toid, source_network$id)
outlet_value <- get_outlet_value(restored)
detoid_idx <- which(restored$toid == outlet_value)
restored$toid[detoid_idx] <- src_lookup[as.character(restored$id[detoid_idx])]
```

The domain's segment is now a connected sub-basin again; standard
hydroloom operations (model coupling, hydraulic routing, end-to-end
visualization) work unchanged. Switching back is a one-column
assignment.

# Functional characteristics

The decomposition operates on drainage basins. A drainage basin is the
total upstream area draining to an outlet, with a single mainstem
connecting its headwater to that outlet. The input network may contain
one or more drainage basins, each composed of incremental catchments
connected in an upstream-downstream topology.

For each drainage basin, `decompose_network()` identifies the
basin's extensive network — the outlet levelpath by default, or the
union of levelpaths whose outlet metric exceeds `stem_threshold` — and
divides it into segments at confluences. Each segment becomes a domain
whose catchments slot holds the lateral tributaries draining into the
segment together with the segment's extensive network rows in their
decomposed form. Even segments with no lateral tributaries become
domains — the extensive network rows alone constitute the segment's
contributing area.

The basin's extensive network overlay is built alongside the domains.
It is a `hy_leveled` view of the same connecting flowlines, with
`toid`s intact, so the connected mainstem is addressable end-to-end.
The same rows appear in their decomposed form across the segment
domains.

A domain corresponds to an HY_Features catchment aggregate. Its
drainage area is the full contributing area to its segment. The
decomposition does not materialize catchment aggregate objects
directly; the alignment is by construction, with the domain's
catchments slot holding the rows the aggregate would.

The inter-domain graph records flow between domains through synthetic
nexus identifiers, with one edge per inter-segment handoff. The graph
is derived on demand from `nexus_registry`; see `get_domain_graph()`.

# Separation of concerns

The decomposition separates two concerns:

1. **Per-domain computation.** Each domain's catchment network is
   self-contained. Processing a domain requires no information from
   any other domain beyond what arrives at inlet nexuses.
2. **Inter-domain connectivity.** The derived domain graph captures
   which domains flow into which, and at what nexus — a
   coarser-resolution network that can be traversed independently of
   catchment-level detail.

# Data model

The decomposition produces a `domain_decomposition` object with six
slots:

- **`domains`** — named list of `hy_domain` objects, one per
  sub-basin segment.
- **`domain_connectivity`** — named list of `hy_leveled` overlays
  keyed by basin id. Each overlay is the basin's *extensive
  network*: a `hy_leveled` view of the connecting flowlines with
  `toid`s intact except at the basin outlet, which carries the
  reserved outlet `toid` value. Sub-threshold basins have no overlay.
- **`overrides`** — non-dendritic inter-domain transfer table, or
  `NULL`.
- **`catchment_domain_index`** — named character vector mapping each
  catchment id to its domain id.
- **`nexus_registry`** — synthetic nexus identifiers and the domains
  they connect.
- **`source_network`** — the original enriched input network.

The inter-domain edge list is not stored; `get_domain_graph()` derives
it from `nexus_registry` on demand.

## `hy_domain`

A `hy_domain` bundles a domain's catchment network with the
connectivity metadata that recomposition needs:

- `domain_id` — unique identifier.
- `outlet_nexus_id` — the hydro nexus where this domain discharges.
- `inlet_nexus_ids` — hydro nexus ids where upstream domains feed
  into this one. `character(0)` for leaf domains; populated for stem
  and root domains.
- `containing_domain_id` — the enclosing domain for endorheic
  basins, or `NA`.
- `catchments` — hydroloom object carrying the domain's catchment
  network. Catchments include both lateral tributary rows and the
  segment's extensive network rows; the extensive network rows have
  a `toid` set to `get_outlet_value()`. To tell the two apart,
  intersect the domain's `catchments$id` with the basin's extensive
  network overlay's `id`.

The `catchments` slot accepts `hy_topo`, `hy_leveled`, or
`hy_flownetwork`.

# S3 classes

The decomposition builds on the hydroloom S3 class hierarchy described
in `vignette("hydroloom")`. The relevant classes are:

- **`hy_topo`** — self-referencing edge list (`id`, `toid`; unique
  `id`).
- **`hy_leveled`** — subclass of `hy_topo` enriched with `topo_sort`,
  `levelpath`, and `levelpath_outlet_id`. Required input for
  `decompose_network()`, because the partition uses `levelpath` to
  identify extensive network membership and `topo_sort` to establish
  processing order.
- **`hy_flownetwork`** — non-dendritic edge list (`id`, `toid`;
  non-unique `id` allowed). Domains may hold `hy_flownetwork`
  catchments to preserve internal divergences.

# Running a decomposition

The input must be enriched to `hy_leveled` before decomposition.
`add_toids()` and `add_levelpaths()` handle the topology and
levelpath enrichment; `add_streamorder()` adds the
`stream_calculator` column that `decompose_network()` consults when
`stem_threshold` is supplied (see `Extensive Network selection`
below). For data carrying a `StreamCalc` column from NHDPlus, the
column is already canonicalized by `hy()` and `add_streamorder()`
becomes a no-op; for other sources it is required.

```{r decompose}
library(hydroloom)

g <- sf::read_sf(system.file("extdata/new_hope.gpkg", package = "hydroloom"))

src <- hy(g) |>
  add_toids(return_dendritic = TRUE) |>
  add_levelpaths(name_attribute = "GNIS_ID",
    weight_attribute = "arbolate_sum") |>
  add_streamorder(status = FALSE)

d <- decompose_network(src)
```

The returned `domain_decomposition` prints a summary of the partition:

```{r print}
d
```

## Extensive Network selection

`decompose_network()` determines which catchments carry the basin's
extensive network through one of three paths:

- **Default** (both `stem_threshold` and `stem_levelpaths` are
  `NULL`) — the extensive network is the basin's outlet levelpath.
- **Threshold-based** — when `stem_threshold` is supplied, a basin's
  outlet metric must exceed the threshold for the basin to be split
  along its extensive network. Sub-threshold basins become a single
  domain with no extensive network overlay.
- **Explicit** — when `stem_levelpaths` is supplied, those
  levelpaths (plus the outlet levelpath) form the extensive network.

```{r threshold}
d_thresh <- decompose_network(src, stem_threshold = 100)

d_thresh
```

The map below colors each flowline by the domain it belongs to and
overlays the basin's extensive network in black. Every flowline
belongs to some domain (the partition covers the network); the
overlay highlights the rows that are *also* in the basin's extensive
network — the connecting flowlines that the domains turn into local
outlets.

```{r plot-domains, fig.width = 7, fig.height = 7}
# Build a domain_id column on the source flowlines.
domain_idx <- d_thresh$catchment_domain_index
plot_net <- d_thresh$source_network
plot_net$domain_id <- unname(domain_idx[as.character(plot_net$id)])

# Pull all extensive network ids from every basin's overlay.
conn_ids <- unlist(lapply(d_thresh$domain_connectivity,
  \(o) as.character(o$id)),
  use.names = FALSE)

domain_ids <- sort(unique(plot_net$domain_id))

set.seed(42)
pal <- sample(grDevices::hcl.colors(length(domain_ids), palette = "Set 2"))
names(pal) <- domain_ids

par(mar = c(0, 0, 1, 0))
plot(sf::st_geometry(plot_net),
  col = pal[plot_net$domain_id],
  lwd = 2, main = "Domain decomposition (stem_threshold = 100)")
plot(sf::st_geometry(plot_net[as.character(plot_net$id) %in% conn_ids, ]),
  col = "black", lwd = 2, add = TRUE)
legend("topright", legend = c("extensive network overlay",
  "domain (each color = one)"),
  col = c("black", "grey60"), lwd = 2, bty = "n", cex = 0.8)
```

## Manual domain breaks

The automatic segmentation splits at confluences —
every point where a tributary enters starts a new
segment. A large threshold pulls more rivers into domains 
producing fewer, larger domains. The `domain_breaks` parameter
gives direct control over where the network is segmented,
independent of the confluence structure.

In practice, this is useful when domain boundaries need to align with
modeling sub-basins, gauge locations, or administrative units rather
than the topology's natural confluence points. The approach is to set
a high threshold and
then pass specific catchment ids as break points.

```{r manual-breaks}
# A high threshold pulls many levelpaths into the extensive network.
# Without domain_breaks the extensive network is segmented at every
# confluence, which may produce more domains than desired.
d_high <- decompose_network(src, stem_threshold = 25)
d_high

# Identify extensive network catchments where we want to force segment
# boundaries. Here we pick two catchments roughly in the middle of
# the extensive network overlay.
conn <- d_high$domain_connectivity[[1]]

# Choose break points at the 1/3 and 2/3 marks along the extensive
# network.
break_ids <- conn$id[
  c(floor(nrow(conn) / 3), floor(2 * nrow(conn) / 3))]

d_manual <- decompose_network(src,
  stem_threshold = 25,
  domain_breaks = break_ids)
d_manual
```

The map below compares the automatic segmentation (left) with the
manual-break segmentation (right). Both use the same threshold; the
difference is where domains attach along the extensive network.

```{r plot-manual-breaks, fig.width = 10, fig.height = 5}
plot_decomp <- function(d, title) {

  domain_idx <- d$catchment_domain_index
  net <- d$source_network
  net$domain_id <- unname(domain_idx[as.character(net$id)])

  conn_ids <- unlist(lapply(d$domain_connectivity,
    \(o) as.character(o$id)),
    use.names = FALSE)

  domain_ids <- sort(unique(net$domain_id))

  set.seed(42)
  pal <- sample(grDevices::hcl.colors(
    max(length(domain_ids), 1), palette = "Set 2"))
  names(pal) <- domain_ids

  plot(sf::st_geometry(net),
    col = pal[net$domain_id],
    lwd = 2, main = title)
  plot(sf::st_geometry(net[as.character(net$id) %in% conn_ids, ]),
    col = "black", lwd = 2, add = TRUE)

}

par(mfrow = c(1, 2), mar = c(0, 0, 2, 0))
plot_decomp(d_high, "Automatic segmentation")
plot_decomp(d_manual, "Manual domain_breaks")
```

## Recomposition

`recompose(decomposition, var)` accumulates a numeric variable on
`source_network` through the decomposed fabric and returns the source
network with the per-row downstream-accumulated value populated.

The function runs in two passes. Pass 1 calls
`accumulate_downstream()` on each domain — the lateral rows roll up
within the domain, and each extensive-network row picks up its
locally-incremental value (own area plus the laterals draining into
it). Pass 2 walks each basin's extensive-network overlay end-to-end,
seeded with the pass-1 values, and produces the basin-wide cumulative
at every extensive-network row.

```{r recompose}
rec <- recompose(d, "da_sqkm")

head(rec[, c("id", "da_sqkm")])
```

Recomposing through the decomposed fabric returns the same values
that `accumulate_downstream()` would produce on the un-decomposed
network, within float-join tolerance:

```{r recompose-check}
reference <- accumulate_downstream(src, "da_sqkm", quiet = TRUE)

ord <- match(src$id, rec$id)

max(abs(rec$da_sqkm[ord] - reference))
```

The decomposed pathway is what makes per-domain parallelism worthwhile:
each domain's pass-1 work is independent, so the expensive accumulation
splits cleanly across cores, and the basin-overlay pass is cheap.

## Containment

`decompose_network()` does not detect containment. Disconnected
components in the input become independent basins — endorheic basins,
drainage-divide remnants, and any other isolated component are
partitioned into their own domains and processed in parallel like any
other basin.

A domain corresponds to a catchment aggregate in HY_Features terms:
a set of catchments grouped together that is itself a catchment, with
inflows, outflows, and a divide. Declaring containment with
`set_containment()` says that one catchment aggregate (the contained
domain) is enclosed by another (the containing domain). The
declaration is typically made because the two share a divide rather
than because flow crosses between them — no hydro nexus passes water
from the contained basin to the containing one.

The relationship is recorded on the contained domain's
`containing_domain_id` slot, surfaced by
`get_domain_graph(relations = "contained")`, and applied by
`recompose()` when called with `containment = "accumulate"`. In that
mode, the contained basin's accumulated value at its outlet is added
at the containing domain's outlet — the most-downstream row of the
containing domain's segment of the extensive network — and routed
downstream from there through the containing basin's extensive
network.

The default `containment = "ignore"` leaves each basin's accumulated
value at its own outlet, which is correct for a true endorheic basin.
The opt-in `accumulate` mode is a modeling choice — useful when a
contained basin's contribution should roll into its container, not a
claim about physical flow across the divide. Containment is
transitive: A inside B inside C accumulates correctly because
recompose works from the most-contained outward, so A's contribution
is added to B before B's value is read, and B's combined value is
added to C.

## Accessing domains

Individual domains, the inter-domain graph, the connectivity
overlays, and the catchment-to-domain mapping are accessible through
dedicated accessors:

```{r access}
# look up a domain by id
get_domain(d, names(d$domains)[[1]])$domain_id

# classify domains as leaves, stems, or roots
domain_ids <- names(d$domains)
roles <- vapply(domain_ids, function(id) {
  if (is_root_domain(d, id)) "root"
  else if (is_leaf_domain(d, id)) "leaf"
  else "stem"
}, character(1))
table(roles)

# inter-domain flow edges (derived from nexus_registry)
head(get_domain_graph(d, relations = "flow"))

# look up which domain owns a catchment
get_domain_for_catchment(d, d$source_network$id[1:5])
```

```{r teardown, include = FALSE}
options(oldoption)
```
