---
title: "Hydroloom Overview"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-dendritic networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.width=4,
  fig.height=4,
  fig.align = "center"
)

library(sf)

oldoption <- options(scipen = 9999)
```

# `hy` S3 class

Hydroloom implements an `hy` S3 class designed to simplify integration of `hydroloom` with existing data and workflows. `hy()` converts a data.frame to an `hy` `data.frame` with attributes compatible with `hydroloom` functions. `hy_reverse()` converts a `hy` `data.frame` back to its original attribute names. You can teach `hydroloom` how to map your attributes to `hydroloom_name_definitions()` with the `hydroloom_names()` function.8  

Most `hydroloom` functions will work with either a `hy` object or a `data.frame` containing names registered with `hydroloom_names()`. Any attributes added to the `data.frame` will contain names from `hydroloom` and must be renamed in the calling environment. 

Internally, the `hy` S3 class has an attribute `orig_names` as shown below. The `orig_names` attribute is used to convert original attribute names back to their original values. Using the `hydroloom` names and the `hy` S3 object are note required but adopting `hydroloom_names_definitions()` may be helpful for people aiming for consistent, simple, and accurate attribute names.


```{r}
library(hydroloom)

x <- sf::read_sf(system.file("extdata/new_hope.gpkg", package = "hydroloom")) |>
  dplyr::select(COMID, REACHCODE, FromNode, ToNode, Hydroseq, TerminalFl, Divergence)

hy(x[1:3,])

attr(hy(x), "orig_names")

```
# Network Graph Representation

The `make_index_ids()` `hydroloom` function creates an adjacency matrix representation of a flow network as well as some conveniet content that are useful when traversing the graph. This adjacency matrix is used heavily in `hydroloom` functions but and may be useful for people who want to write their own graph traversal algorithms.

A companion to `make_index_ids()`, `make_fromids()` reverses the graph such that the adjacency matrix is directed opposite the direction of the provided graph. This is used for upstream flow network traversals.

In the example below we'll add a dendritic toid and explore the `make_index_ids()` output.

```{r}
y <- add_toids(x, return_dendritic = TRUE)

ind_id <- make_index_ids(y)

names(ind_id)

dim(ind_id$to)

max(lengths(ind_id$lengths))

names(ind_id$to_list)

sapply(ind_id, class)
```

Now we'll look at the same thing but for a non dendritic set of toids. Notice that the `to` element of `ind_id` now has three rows. This indicates that one or more of the connections in the matrix has three downstream neighbors. The `lengths` element indicates how many non `NA` values are in each column of the matrix in the `to` element.  


```{r}

y <- add_toids(drop_geometry(x), return_dendritic = FALSE)

ind_id <- make_index_ids(y)

names(ind_id)
dim(ind_id$to)

max(ind_id$lengths)

sum(ind_id$lengths == 2)
sum(ind_id$lengths == 3)

names(ind_id$to_list)

sapply(ind_id, class)
```

To be continued...
