---
title: "Non-dendritic networks"
author: "dblodgett@usgs.gov"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-dendritic networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
   .main-container {max-width: 100%;}
   .row {display: flex;}
   .column {flex: 50%;}
</style>

```{r setup, include = FALSE}
library(hydroloom)
library(dplyr)

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.width=4,
  fig.height=4,
  fig.align = "center"
)

oldoption <- options(scipen = 9999)

```

# Introduction 

`vignette("advanced_network")` talks about the basics of network topology representation and attributes that build on a strictly dendritic network. This vignette expands those topics by describing `hydroloom` functionality that supports non-dendritic networks.

# Background

As discussed in the `vignette("advanced_network")` vignette, a hydrologic flow network can be represented as an edge to edge (e.g. edge list) topology or an edge-node topology. In the case of dendritic networks, an edge list can be stored as a single "toid" attribute on each feature and nodes are redundant as there would be one and only one node for each feature. In non-dendritic networks, an edge list can include multiple "toid" attributes for each feature, necessitating a one to many relationship that can be difficult to interpret. Nevertheless, the U.S. National Hydrography Dataset uses an edge-list format in its "flow table" and the format is capable of storing non-dendritic feature topology. 

Using a node topology to store a flow network, each feature flows from one and only one node and flows to one and only one node. This one to one relationship between features and their from and to nodes means that the topology fits in a table with one row per feature as is common practice in spatial feature data. 

For this reason, the NHDPlus data model converts the NHD "flow table" into node topology in its representation of non dendritic topology. The downside of this approach is that it requires creation of a node identifier. These node identifiers are a table deduplication device that enables a one to many relationship (the flow table) to be represented as two one to one relationships. Given this, in hydrologic flow networks, node identifiers can be created based on an edge list and discarded when no longer needed. 

<div class = "row">
<div class = "column">

```{r node, fig.show="hold", fig.width=3, out.width="45%", echo=FALSE, eval=TRUE}
x <- c(2, 2, 3, 2, 2)
y <- c(5, 4, 3, 2, 1)

a <- c(1.4, 3.5)
b <- c(.9, 5.1)

main_col = "darkblue"
div_col = "purple"

par(mar = c(0, 0, 0, 0))
plot(a, b, col = NA)
points(x, y)
text(c(2.15, 1.85, 3.02, 1.85, 2.15), c(5, 4, 3.2, 2, 1), c("N1", "N2", "N3", "N4", "N5"))

make_edges <- function() {
  arrows(x[1], y[1] - .1, x[2], y[2] + .1, length = .1, col = main_col)
  arrows(x[2] + .1, y[2] - .1, x[3] - .1, y[3] + .1, length = .1, col = div_col) # right
  arrows(x[2] + .0, y[2] - .1, x[4] - .0, y[4] + .1, length = .1, col = main_col)
  arrows(x[3] - .1, y[3] - .1, x[4] + .1, y[4] + .1, length = .1, col = div_col)
  arrows(x[4], y[4] - .1, x[5], y[5] + .1, length = .1, col = main_col)
  text(c(2.1, 2.5, 2.5, 1.9, 2.1), c(4.5, 3.8, 2.3, 3, 1.5), c("1", "2", "3", "4", "5"))
}

make_edges()

```

</div>

<div class = "column">

```{r node2, fig.show="hold", fig.width=3, out.width="45%", echo=FALSE, eval=TRUE}

par(mar = c(0, 0, 0, 0))
plot(a, b, col = NA)

make_edges()

# 
# par(mar = c(0, 0, 0, 0))
# plot(x, y)
# arrows(x[1] + 0.1, y[1] - 0.1, x[3] - 0.1, y [3] + 0.1, 0.1)
# arrows(x[2] - 0.1, y[2] -0.1, x[3] + 0.1, y [3] + 0.1, 0.1)
# arrows(x[3], y[3] - 0.1, x[4], y [4] + 0.1, 0.1)
# text(c(2, 4, 3.1), c(4.2, 4.2, 2), c("1", "2", "3"))
# text(c(1, 5, 3, 3.25), c(4.8, 4.8, 3.4, 1), c("N1", "N2", "N3", "N4"))
```
</div>
</div>

In this example of an edge list topology and a node topology for the same system, feature '1' flows to two edges but only one node. We can represent this in tabular form with a duplicated row for the divergence downstream of '1' or with the addition of node identifiers as shown in the following tables.

<div class = "row">
<div class = "column">

| id | fromnode | tonode |
|----|----------|--------|
| 1 | N1 | N2 |
| 2 | N2 | N3 |
| 3 | N3 | N4 |
| 4 | N2 | N4 |
| 5 | N4 | N5 |

</div>

<div class = "column">

| id | toid |
|----|------|
| 1 | 2 |
| 1 | 4 | 
| 2 | 3 |
| 3 | 5 | 
| 4 | 5 | 
| 5 | 0 | 

</div>
</div>

# Non-dendritic topology attributes. 

Non-dendritic networks represent downstream diverted flow where one path is primary and all others are thought to be secondary. The following attributes are supported by `hydroloom` to help track and work with this primary and secondary downstream categorization.

## fromnode and tonode

The attributes `fromnode` and `tonode` are used to store a flow network as a edge node topology where every network feature has one and only one node upstream and one and only one node downstream. Nodes are useful if converting a flow network to a graph and are useful in many analysis contexts as there is a single identifier for a confluence or divergence.

## divergence

The `divergence` attribute indicates if a downstream connection is primary (1) or secondary (2). If 0, a connection is not downstream of a divergence. This attribute is useful as it facilitates following a flow network in the "downstream mainstem" direction at every divergence.

## return divergence

The `return divergence` attribute indicates that one or more of the features upstream of a given feature originates from a divergence.  If 0, the upstream features are not part of a diversion. If 1, one or more of the upstream features is part of a diversion. 

## stream calculator

The `stream calculator` attribute is part of the modified Strahler stream order as implemented in the NHDPlus data model. It indicates if a given feature is part of the downstream mainstem dendritic network or is part of a diverted path. If 0, the path is part of a diversion. Otherwise `stream calculator` will be equal to stream order. When generating Strahler stream order, if stream calculator is 0 for a given feature, that feature is not considered for incrementing downstream stream order.  

## summary

As a system, `stream calculator`, `divergence` and `return divergence` support network navigation and processing in the context of diverted paths.

1. A feature at the top of a diversion will have `divergence` set to 1.
1. All features that are part of a diversion that has not yet recombined with a main path, will have `stream calculator` set to 0.
1. A feature that is just downstream of where a diversion path recombines with a main path will have `return diversion` set to 1. 
